Exploratory data analysis of DSRCT Samples

### 2024-11-25 analyses/cell-type-dsrct/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
fix typos and re-render notebook

---
title: "EDA of DSRCT Samples"
author: Danh Truong
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "2024-07-08")
#sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)

@@

```{r}
metadata_file <- file.path(data_dir, "SCPCP000013", "single_cell_metadata.tsv")
metadata <- read.csv(metadata_file,  sep = '\t')
metadata_DSRCT <- dplyr::filter(metadata, diagnosis == 'Desmoplastic small round cell tumor')

```

@@

#classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
#output_classifications_file <- file.path(results_dir, classifications_filename)
```

@@

#adding the sample id to the name of each SingleCellExperiment
names(sce_list) <- metadata_DSRCT$scpca_library_id

@@

```{r}
umap_df_list <- lapply(sce_list, function(x) {
# pull out the UMAP coordinates and genes and make a data frame to use for plotting
  umap_df <- x |>
    scuttle::makePerCellDF(features = marker_genes, use.dimred = "UMAP") |>
    # replace UMAP.1 with UMAP1
    dplyr::rename_with(\(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")) |>
    # combine all genes into a single column for easy faceting
    tidyr::pivot_longer(cols = starts_with("ENSG"),
                        names_to = "ensembl_gene_id",
                        values_to = "gene_expression") |>
    # join with marker gene df to get gene symbols for plotting
    dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
    dplyr::select(barcodes,
                  UMAP1,
                  UMAP2,
                  gene_symbol,
                  ensembl_gene_id,
                  gene_expression,
                  cluster)
})

```

```{r, fig.width=8}
for(i in 1:length(umap_df_list)){
  # faceted umap showing a umap panel for each marker gene
  p <- ggplot(umap_df_list[[i]], aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.8, size = 0.1) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(color = "Log-normalized gene expression") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
    ggtitle(names(umap_df_list)[i])

  #save plots
  ggsave(
    filename = paste0(
      '../plots/marker_expression_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )
}

```

In my experience, ST6GALNAC5 is a strong marker for DSRCT in single-cell data. As can be seen, several of the samples have expression of ST6GALNAC5, but some do not, such as SCPCS000731 and SCPCS000729. In fact, these SCPCS000729 contains  low number of cells. These  samples in particular are the PDX samples and I will be excluding SCPCS000729 from further analyses.

```{r}
umap_df_list_excluded <- umap_df_list[!(names(umap_df_list) %in% c('SCPCS000731'))]
```

We can also look at the distributions for each marker gene. 
I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene. What is clear from these plots that ST6GALNAC5, CACNA2D2, PTPRQ, IQCJ-SCHIP1,  show a bi modal distribution among the cells. To some extent, we do see expression of the other markers.

```{r}
for(i in 1:length(umap_df_list_excluded)) {
  p <- ggplot(umap_df_list_excluded[[i]],
              aes(x = gene_expression, fill = gene_symbol)) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = 'free_y') +
    theme(legend.position = "none") + 
    ggtitle(names(umap_df_list_excluded)[i])
  print(p)

  #save plots
  ggsave(
    filename = paste0(
      '../plots/marker_distribution_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )
}
```

@@

```{r}
umap_df_list_excluded_scaled <- lapply(umap_df_list_excluded, function(x){
  x   |>
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
  dplyr::ungroup()
})

```

Now we can create the same density plot but looking at z-scores. Interestingly, in some samples, like SPCS000489, we see that ST6GALNAC5 has negative values. Presumably, these are tumor cells but there may be a skew due to the number of tumor cells compared to normal.

```{r}

for(i in 1:length(umap_df_list_excluded_scaled)) {
  p <- ggplot(umap_df_list_excluded_scaled[[i]],
              aes(x = transformed_gene_expression, fill = gene_symbol)) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = 'free_y') +
    theme(legend.position = "none") + 
    ggtitle(names(umap_df_list_excluded_scaled)[i])

@@

  #save plots
  ggsave(
    filename = paste0(
      '../plots/marker_distribution_zscaled_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )

}
```

@@

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
  x |>
  dplyr::group_by(barcodes) |> 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()
})

@@

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  #save plots
  ggsave(
    filename = paste0(
      module_base,
      '/plots/UMAP_marker_expression_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )
}

```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

@@

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- lapply(marker_sum_exp, function(x){
  x |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 1) +
  ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  #save plots
  ggsave(
    filename = paste0(
      module_base,
      '/plots/UMAP_classification_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )
}

```

@@

```{r}
# calculate sum gene expression across all marker genes in list 
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x){
  x |> 
  dplyr::group_by(cluster) |> #change to cluster
  dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |> 
  dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |> 
  dplyr::distinct()
})

@@

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  #save plots
  ggsave(
    filename = paste0(module_base, '/plots/UMAP_marker_expression_cluster_', names(umap_df_list)[i], '.png'),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )
}

```
It looks like there are group of cells with high marker gene expression. We  anticipate that these are most likely to be the tumor cells. The groups with low or negative values are likely normal cells.

@@

```{r}
# classify tumor cells based on presence of any marker genes 
marker_sum_exp <- lapply(marker_sum_exp, function(x){
  d <- density(x$sum_exp)

  x |> 
  dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))})

for(i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 1) +
  ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  #save plots
  ggsave(
    filename = paste0(
      module_base,
      '../plots/UMAP_classification_cluster_',
      names(umap_df_list)[i],
      '.png'
    ),
    plot = p,
    width = 6,
    height = 6,
    units = 'in',
    dpi = 150
  )

}

```

@@

## Save outputs 
```{r}
# get an RDS of the processed data
saveRDS(sce_file_list, '../results/SCPCP000013_sce_file_list.rds')
```

>>>>

---
title: "Exploratory data analysis of DSRCT Samples"
author: Danh Truong
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current")
# sample_dir <- file.path(data_dir, "SCPCP000013", params$sample_id)

@@

```{r}
metadata_file <- file.path(data_dir, "SCPCP000013", "single_cell_metadata.tsv")
metadata <- read.csv(metadata_file, sep = "\t")
metadata_DSRCT <- dplyr::filter(metadata, diagnosis == "Desmoplastic small round cell tumor")
```

@@

# classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
# output_classifications_file <- file.path(results_dir, classifications_filename)
```

@@

# adding the sample id to the name of each SingleCellExperiment
names(sce_list) <- metadata_DSRCT$scpca_library_id

@@

```{r}
umap_df_list <- lapply(sce_list, function(x) {
  # pull out the UMAP coordinates and genes and make a data frame to use for plotting
  umap_df <- x |>
    scuttle::makePerCellDF(features = marker_genes, use.dimred = "UMAP") |>
    # replace UMAP.1 with UMAP1
    dplyr::rename_with(\(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")) |>
    # combine all genes into a single column for easy faceting
    tidyr::pivot_longer(
      cols = starts_with("ENSG"),
      names_to = "ensembl_gene_id",
      values_to = "gene_expression"
    ) |>
    # join with marker gene df to get gene symbols for plotting
    dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
    dplyr::select(
      barcodes,
      UMAP1,
      UMAP2,
      gene_symbol,
      ensembl_gene_id,
      gene_expression,
      cluster
    )
})
```

```{r, fig.width=8}
for (i in 1:length(umap_df_list)) {
  # faceted umap showing a umap panel for each marker gene
  p <- ggplot(umap_df_list[[i]], aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.8, size = 0.1) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(color = "Log-normalized gene expression") +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
    ggtitle(names(umap_df_list)[i])

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/marker_expression_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```

In my experience, `ST6GALNAC5` is a strong marker for DSRCT in single-cell data. As can be seen, several of the samples have expression of `ST6GALNAC5`, but some do not, such as SCPCS000731 and SCPCS000729. In fact, these SCPCS000729 contains  low number of cells. These samples in particular are the PDX samples and I will be excluding SCPCS000729 from further analyses.

```{r}
umap_df_list_excluded <- umap_df_list[!(names(umap_df_list) %in% c("SCPCS000731"))]
```

We can also look at the distributions for each marker gene. 
I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene. What is clear from these plots that `ST6GALNAC5`, `CACNA2D2`, `PTPRQ`, `IQCJ-SCHIP1`,  show a bi modal distribution among the cells. To some extent, we do see expression of the other markers.

```{r}
for (i in 1:length(umap_df_list_excluded)) {
  p <- ggplot(
    umap_df_list_excluded[[i]],
    aes(x = gene_expression, fill = gene_symbol)
  ) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = "free_y") +
    theme(legend.position = "none") +
    ggtitle(names(umap_df_list_excluded)[i])
  print(p)

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/marker_distribution_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```

@@

```{r}
umap_df_list_excluded_scaled <- lapply(umap_df_list_excluded, function(x) {
  x |>
    dplyr::group_by(gene_symbol) |>
    # get z-scores for each gene
    dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
    dplyr::ungroup()
})
```

Now we can create the same density plot but looking at z-scores. Interestingly, in some samples, like SCPCS000489, we see that `ST6GALNAC5` has negative values. Presumably, these are tumor cells but there may be a skew due to the number of tumor cells compared to normal.

```{r}
for (i in 1:length(umap_df_list_excluded_scaled)) {
  p <- ggplot(
    umap_df_list_excluded_scaled[[i]],
    aes(x = transformed_gene_expression, fill = gene_symbol)
  ) +
    geom_density() +
    facet_wrap(vars(gene_symbol), scales = "free_y") +
    theme(legend.position = "none") +
    ggtitle(names(umap_df_list_excluded_scaled)[i])

@@

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/marker_distribution_zscaled_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```

@@

```{r}
# calculate sum gene expression across all marker genes in list
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x) {
  x |>
    dplyr::group_by(barcodes) |>
    dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |>
    dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |>
    dplyr::distinct()
})

@@

for (i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
    geom_point(size = 0.5, alpha = 0.5) +
    scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/UMAP_marker_expression_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

@@

```{r}
# classify tumor cells based on presence of any marker genes
marker_sum_exp <- lapply(marker_sum_exp, function(x) {
  x |>
    dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))
})

for (i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
    geom_point(size = 0.5, alpha = 1) +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/UMAP_classification_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```

@@

```{r}
# calculate sum gene expression across all marker genes in list
marker_sum_exp <- lapply(umap_df_list_excluded_scaled, function(x) {
  x |>
    dplyr::group_by(cluster) |> # change to cluster
    dplyr::mutate(sum_exp = sum(transformed_gene_expression, na.rm = T)) |>
    dplyr::select(barcodes, UMAP1, UMAP2, sum_exp, cluster) |>
    dplyr::distinct()
})

@@

for (i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
    geom_point(size = 0.5, alpha = 0.5) +
    scale_color_viridis_c() +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  # save plots
  ggsave(
    filename = paste0(module_base, "/plots/UMAP_marker_expression_cluster_", names(umap_df_list)[i], ".png"),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```
It looks like there are group of cells with high marker gene expression. We  anticipate that these are most likely to be the tumor cells. The groups with low or negative values are likely normal cells.

@@

```{r}
# classify tumor cells based on presence of any marker genes
marker_sum_exp <- lapply(marker_sum_exp, function(x) {
  d <- density(x$sum_exp)

  x |>
    dplyr::mutate(sum_classification = dplyr::if_else(sum_exp > 0, "Tumor", "Normal"))
})

for (i in 1:length(marker_sum_exp)) {
  p <- ggplot(marker_sum_exp[[i]], aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
    geom_point(size = 0.5, alpha = 1) +
    ggtitle(names(umap_df_list_excluded_scaled)[i])
  print(p)

  # save plots
  ggsave(
    filename = paste0(
      module_base,
      "/plots/UMAP_classification_cluster_",
      names(umap_df_list)[i],
      ".png"
    ),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 150
  )
}
```

@@

## Save outputs 
```{r}
# get an RDS of the processed data
saveRDS(sce_file_list, "../results/SCPCP000013_sce_file_list.rds")
# get an RDS of the UMAP data with marker genes
saveRDS(umap_df_list, file.path(module_base, "results/SCPCP000013_umap_df_list.rds"))
```
